### 2023-03-02 src/explore.Rmd
clean

---
title: "Pretty darn good control: when are approximate solutions better than approximate models"
titlerunning: Pretty darn good control
thanks: | 
    This material is based upon work supported by the National Science
    Foundation under Grant No. DBI-1942280.

authors: 
- name: Felipe Montealegre-Mora
  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
- name: Marcus Lapeyrolerie
  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
- name: Melissa Chapman
  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
- name: Abigail G. Keller
  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
- name: Carl Boettiger
  address: Department of Environmental Science, Policy, and Management, University of California, Berkeley
  email: cboettig@berkeley.edu

keywords:
- Optimal Control
- Reinforcement Learning
- Uncertainty
- Decision Theory

#PACS: 
#- PAC1
#- superPAC

#MSC:
#- MSC code 1
#- MSC code 2

abstract: |
  Existing methods for optimal control struggle to deal with the
  complexity commonly encountered in real-world systems, including
  dimensionality, process error, model bias and data heterogeneity.
  Instead of tackling these system complexities directly, researchers 
  have typically sought to simplify models to fit optimal control 
  methods. But when is the optimal solution to an approximate, stylized 
  model better than an approximate solution to a more accurate model? 
  While this question has largely gone unanswered owing to the 
  difficulty of finding even approximate solutions for complex models, 
  recent algorithmic and computational advances in deep reinforcement 
  learning (DRL) might finally allow us to address these questions. DRL 
  methods have to date been applied primarily in the context of games 
  or robotic mechanics, which operate under precisely known rules. Here, 
  we demonstrate the ability for DRL algorithms using deep neural networks 
  to successfully approximate solutions (the ‘‘policy function’’ or control 
  rule) in a non-linear three-variable model for a fishery without knowing 
  or ever attempting to infer a model for the process itself. We find that 
  the reinforcement learning agent discovers an effective simplification of
  the problem to obtain an interpretable control rule. We show that the 
  policy obtained with DRL is both more profitable and more sustainable 
  than any constant mortality policy---the standard family of policies 
  considered in fishery management.

bibliography: bibliography.bib
biblio-style: spphys

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \usepackage{setspace}
  - \usepackage{bookmark}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \usepackage{tikz-cd}
  - \usepackage{tcolorbox}
  - \usepackage{etoolbox}
  - \usepackage{url}
  - \linenumbers
  - \doublespacing

# bibstyle options spbasic(default), spphys, spmpsci
# output: rticles::springer_article

output: 
  rticles::arxiv_article:
    keep_tex: true

editor_options: 
  markdown: 
    wrap: sentence

---

```{r knit_global, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, 
                      warning = FALSE, fig.height=5, fig.width=7.5)
library(tidyverse)
library(patchwork)
```

```{r}
# for best const mortality and best const escapement
msy_optimizer <- function(df) {
  df |> 
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}
escapement_optimizer <- function(df) {
  df |> 
  group_by(escapement, rep) |>
  filter(t == max(t)) |> 
  group_by(escapement) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}

# Collect data (outsource to separate R script later TBD)
msy_df <- read_csv("../data/msy.csv.xz")
best_action <- msy_optimizer(msy_df)$action
escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
best_e <- escapement_optimizer(escapement_df)$escapement
ppo_df <- read_csv("../data/PPO200.csv.xz")

opt_escapement <- escapement_df |> filter(escapement == best_e)

# set up for plots:

@@

<!--
 Special Issue Description:

This special issue highlights the development of novel data-driven methods, including statistics, machine learning, parameter estimation, and uncertainty quantification, and combinations thereof, towards modeling biological systems. These newly developed methods will tackle challenges that are commonly encountered when modeling real-world experimental, field, pre-clinical, or clinical data. Examples of such challenges include high dimensionality, computational complexity, observation or process error, model bias, and intra- or inter-individual heterogeneity. Contributions to this special issue require validation of new methods with real-world data or simulated data sets that contain features of real-world data that exemplify an outlined modeling challenge. Papers should include a discussion justifying why the developed method is novel and not an application of previously developed methods, as well as how the method may be broadly applicable across different areas of biology, including medical, ecological, genetics, and epidemiological applications.
-->

# Introduction {#intro}

Much effort has been spent grappling with the complexity of our natural world in contrast to the relative simplicity of the models we use to understand it.
Heroic amounts of data and computation are being brought to bear on developing better, more realistic models of our environments and ecosystems, in hopes of improving our capacity to address the many planetary crises.
 But despite these efforts and advances, we remain faced with the difficult task of figuring out how best to respond to these crises.
While simplified models have historically allowed for exploration of large decision spaces, the new wave of rich models are applied to highly oversimplified descriptions of potential actions they seek to inform.
For instance, Global Circulation Models (GCMs) such as HadCM3 [@pope2000impact; @gordon2000simulation; @collins2001internal] model earth's climate using 1.5M variables, while the comparably vast potential action space is modeled as one of six emissions scenarios from the IPCC [@riahi2017shared].

Even as our research community develops simulations of the natural world that fit only in supercomputers, we analyze a space of policies that would fit on index cards.
Similar combinations of rich process models and highly simplified decision models (often not even given the status of 'model') are common.
Modeling the potential action space as one of a handful of discrete scenarios is sometimes a well justified acknowledgement of the constraints faced by real-world decision-makers – particularly in the context of multilateral decisions –
and may seem to reflect a division of responsibilities between 'scientists' modeling the 'natural processes' and policy-makers who make the decisions.
But, more often, this simplification of decision choices is simply mathematically or conceptually convenient.  <!-- there are questions of policy scale hiding here -- e.g. local regulators of a specific industry obviously face a different action space than national legislatures or international diplomats -->
This simplification reflects trade-offs between tractablity and complexity at the basis of any mathematical modeling -- if we make both the state space and action space too realistic, the problem of finding the best sequence of actions quickly becomes intractable.
However, emerging data-driven methods from machine learning offer a new choice -- algorithms that can find good strategies in such problems, but at the cost of opacity.

In this paper, we focus on a well-developed application of model-based management of the natural world that has long illustrated the trade-offs between model complexity and policy complexity: the management of marine fisheries.
Fisheries management is both an important issue to itself and a rich & frequent test-bed of ecological management more generally.
Fisheries are an essential natural resource that provide the primary source of protein for one in every four humans, and have faced widely documented declines due to over-fishing [@Worm2006, @Costello2016].
Fisheries management has  roots in both the fields of _ecosystem management_ and _natural resource economics_.
Both fields might trace their origins to the notion of maximum sustainable yield (MSY), introduced independently by a fisheries ecologist [@Schaefer1954] and the economist [@Gordon1954] in the same year.
From this shared origin, each field would depart from the simplifying assumptions of the Gordon-Schaefer model in divergent ways, leading to different techniques for deriving policies from models.The heart of the management problem is easily understood: a manager seeks to set quotas on fishing that will ensure the long-term profitability and sustainability of the industry.
Mathematical approaches developed over the past century may be roughly divided between these two fields: (A) ecologists, focused on ever more realistic models of the biological processes of growth and recruitment of fish while considering relatively stylized suite of potential management strategies, and (B) economists, focused on far more stylized models of the ecology while exploring a far less constrained set of possible policies.
The economist's approach can be characterized by the mathematics of a Markov decision process [MDP @Clark1973; @Clark1990; @Marescot2013], in which the decision-maker must observe the stock each year and recommend a possible action.
In this approach, the policy space that must be searched is exponentially large -- for a management horizon of T decisions and a space of N actions, the number of possible policies is $N^T$.
In contrast, fisheries ecologists and ecosystem management typically search a space of policies that does not scale with the time horizon.
Under methods such as "Management Strategy Evaluation" (MSE, [@Punt2016]) a manager identifies a candidate set of "strategies" a priori, and then compares the performance of each strategy over a suite of simulations to determine which strategy gives the best outcome (i.e. best expected utility).
This approach is far more amenable to complex simulations of fisheries dynamics and more closely corresponds to how most marine fishing quotas are managed today in the United States [see stock assessments documented in @RAMLegacyDB].
<!-- Do we say here that these strategies retain the 'constant mortality' intuition of @Schaefer? -->

```{r}
#| label: textbox
#| out.width: "1\\linewidth"
knitr::include_graphics("figures/textbox.png")
```

Recent advances in machine learning may allow us to once again bridge these approaches, while also bringing new challenges of their own.
Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RAMLegacyDB].
Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
However, advances _Deep Reinforcement Learning_, (DRL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari1; @atari2] to fusion reactions [@fusion1; @fusion2] to the remarkable dialog abilities of ChatGPT [@chatgpt].
RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, requiring immense computational costs, and presenting frequent challenges with transfer-ability or reproducibility.
A review of all these issues is beyond our scope but can be found elsewhere [@rl-intro; @millie-3].
Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
In general, one can only assess the performance of these black box methods relative to alternatives.

```{r}
#| label: conceptual
#| echo: FALSE
#| message: FALSE
#| fig.align: 'center'
#| fig.cap:  '(A) Classical management strategies are often derived by reducing the complex dynamics of the fishery ecosystem to a single equation governing the harvested population (say, $X$). where the interaction between $X$ and its environment is summarized to a set of parameters, such as the growth rate $r$, and the carrying capacity $K$. An advantage of one-dimensional approaches is that the optimal policy is often known exactly, and, moreover, is intuitive. For example, in the logistic equation shown, the maximal sustainable yield of the system is attained at K/2. 
#| (B) In contrast to this, we study the problem of managing a three species fishery, where we harvest species $X$.
#| In addition to $X$, there is a species $Y$ which competes with $X$ for resources, and a species $Z$ which preys on $X$ and $Y$.
#| We compare two approaches solving harvest strategies in this system. 
#| The first (Harvest $B_{\mathrm{i}}$) is based on the intuition obtained from 1D control: we use simulated data to empirically determine which constant mortality policy is optimal for controlling this fishery.
#| The second (Harvest $B_{\mathrm{ii}}$) is an RL policy, where the policy function is expressed as a neural network and the algorithm uses a variant of \emph{gradient ascent} (namely, \emph{Proximal Policy Optimizaton (PPO)}) to optimize this network.
#| Both approaches we compare are \emph{model-free:} the optimization algorithm treats the model as a black box, only using its input-output statistics.
#| This way, the optimal policy is not computed from the values of the model parameters, but rather by comparing how policies perform in controlling the system.'
#| out.width: '6.5in' 
#| fig.pos: 'H'

knitr::include_graphics("figures/conceptual.png")
```

<!--NEED some kind of map paragraph here saying what's ahead, what we show  -->
In this article, we compare the performance of a DRL solution to that of a management strategy evaluation across potential 'constant mortality' policies.\footnote{A repository with all the relevant code to reproduce our results may be found at \url{https://github.com/boettiger-lab/approx-model-or-approx-soln} in the ‘‘code’’ subdirectory. The data used is found in the ‘‘data’’ subdirectory, but the user may use the code provided to generate new data sets.} 
We consider the problem of devising harvest strategies in a three species ecosystem (Figure 1B).
We show that, somewhat surprisingly, the DRL policy closely resembles a *bang-bang* solution (see Text Box 1).
Bang-bang solutions are known to be optimal for a variety of single-species models, however, there is no a-priori reason why they should be optimal for higher-dimensional systems.
This way, the DRL agent discovers and exploits, as it were, an effective simplification of the control problem.
We provide some intuition as to how this simplification arises from the dynamics of the system, but leave a more rigorous characterization of this phenomenon for future work.

The DRL policy drives an oscillation process in the system and, in this way, maintains the long-term sustainability of fishing.
In contrast, constant mortality policies overfish at low populations and underfish (relative to the DRL solution) at high populations.
This leads to a tradeoff when leveraging constant mortality strategies: higher constant mortality rates lead to high profitability at the expense of a higher probability of overfishing to extinction, while lower constant mortality rates attain stability at the expense of profit.

The DRL approach side-steps this tradeoff by optimizing over a more complex family of possible policies—policies parametrized by a neural network, as opposed to policies labeled by a single parameter (the mortality rate).
We thus observe that the DRL solution leads to a higher profitability *and* stability as compared to constant mortality policies.

# Mathematical models of fisheries

In this section we introduce some models describing the population dynamics of fisheries.
In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
For $n$ species, these models have the general form
\begin{align}
  \label{eq:general model}
  \begin{split}
    \Delta X_t &= f_X(N_t) - M_X X_t\\
    \Delta Y_t &= f_Y(N_t) - M_Y Y_t\\
    &\dots
  \end{split}
\end{align}
where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $\Delta X_t := X_{t+1} - X_t$, $\Delta Y_t := Y_{t+1} - Y_t$, and where $f_i:\mathbb{R}^{n}\to\mathbb{R}$ are arbitrary functions.
Here, $M_i=M_i(N_t)\in\mathbb{R}_+$ is a state-dependent fish mortality arising from harvesting (sometimes referred to as *fishing effort*).
Thus, the term $M_X X_t$ is the total $X$ harvest at time $t$.
We will focus on systems where a single species, $X$, is harvested, i.e. $M_Y = M_Z = \dots = 0$.

## A single species model

Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$, as described in $Text$ $Box$ $1$.
The most familiar model of $f(X)$ is that of *logistic growth*, for which

\begin{align}
  \label{eq:logistic}
  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
\end{align}

Real world ecological systems are obviously far more complicated than this simple model suggests.
One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
A seminal example of such dynamics was introduced in [@may77], using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:

\begin{align}
  \label{eq:may}
  f_{\text{May}}(X_t)
  := L(X_t;\ r, K) - F(X_t, H; \beta,c),
\end{align}
where,
\begin{align*}
  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
\end{align*}

The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal predation rate $\beta$, and the predation rate half-maximum biomass $c$.

Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{fig:may}).
Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $X_t$ of the system, these two attractors separated by an unstable fixed point.
At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
For this value of $H$, and for higher values, only the lower fixed point remains.
This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $X_t$ to the original stable state.

```{r}
#| label: may
#| fig.cap: "The fixed point diagram for the model associated to $f_{\\text{May}}$ as a function of varying the parameter $\\beta H$.
#| That is, these are the zeroes of the function $f_{\\text{May}}$. 
#| Stable fixed points (also known as *attractors*) are plotted using a solid line, while the unstable fixed point is shown as a dotted line."
#| out.width: "6.5in"

knitr::include_graphics("figures/MayStablePoints.pdf")
```

This structure implies two things.
First, that a drift in $H$ could lead to catastrophic consequences, with the population $X_t$ plummeting to the lower fixed stable point.
Second, that if the evolution of $X_t$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.

<!-- Discuss control solutions of this model(?) -->
<!-- Briefly motivate reason to consider 3d model, adding competition and predation processes -->

## A three species model

Consider the following three species generalization of eq. \eqref{eq:may}:

\begin{align}
\label{eq:3d model}
\begin{split}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align}
The model is conceptualized in Fig. \ref{fig:conceptual}, panel B.
It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
There are ten parameters in this model: 
The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
A maximum predation rate $\beta$ and a predation rate half-maximum biomass $c$ specifying how $Z$ forages on $X$ and $Y$.
A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
Finally, a death rate $d_Z$ and a parameter $f$ scaling the birth rate of $Z$.

This model generalizes eq. \eqref{eq:may} in the following sense:
If $f=d_Z=c_{XY}=0$, and we let $Z_{t=0}=H$, then eq. \eqref{eq:3d model} becomes
\begin{align*}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, H;\ \beta, c) - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, H;\ \beta, c),\\
  Z_t &= H.
\end{align*}
The equations for $X_t$ and $Y_t$ thus become decoupled, with $X_t$ evolving according to \eqref{eq:may}.

As can be expected, the dynamics of eq. \eqref{eq:3d model} are significantly more complex than those of eq. \eqref{eq:may}.
There are a variety of fixed points for the equations---around 8 or 9 for wide ranges of parameter values.
Fig. \ref{fig:fixed points} provides a rough idea of the fixed point structure by plotting the bifurcation diagram obtained from varying $D$ and fixing the other parameters.

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stX.png}
    \caption{$X$ component of fixed points.}
    \label{fig:stx}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stY.png}
    \caption{$Y$ component of fixed points.}
    \label{fig:sty}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/stZ.png}
    \caption{$Z$ component of fixed points.}
    \label{fig:stz}
  \end{subfigure}
  \caption{Numerically-computed fixed points of the three-species deterministic model, eq. \eqref{eq:3d model}.
  Here, the parameter $D$ was varied between 0.75 and 1.25, and the values of the other parameters are fixed as in eq. \eqref{eq:param vals}. 
  The three components of the fixed points are plotted separately, as a function of $D$.}
  \label{fig:fixed points}
\end{figure}

Throughout most of this paper we will fix the parameters of the model according to eq. \eqref{eq:param vals}.
At these values, the fixed points of the model are:\footnote{
We have included only three decimal digits for readability.
These fixed points may be generated by running the ‘‘stable\_points.py’’ script in \url{https://github.com/boettiger-lab/approx-model-or-approx-soln}.
}
\begin{align*}
  (X,Y,Z) \approx 
  &(0.839, 0.054, 0.377), \quad 
    (0.048, 0.774, 0.480), \quad
    (0.359, 0.491, 0.509), \\
    &(0.899, 0.000, 0.303), \quad
    (0.000, 0.818, 0.457), \quad
    (0.666, 0.666, 0.000), \\
    &(1.000,0.000,0.000),\quad (0.000,1.000,0.000),\quad (0.000,0.000,0.000).
\end{align*}
The classification of fixed points for multi-dimensional systems is generally more complicated than in one dimension.
A fixed point might be stable against perturbations in one direction, but unstable against perturbations along another.
While a detailed classification of these fixed points is beyond the scope of this work, we point out that the point $(0.839, 0.054, 0.377)$ is stable against perturbations in all directions.
We will thus use it as a starting point for our simulations.

## Stochastic dynamics

We have, up to now, presented eqs. \eqref{eq:may} and \eqref{eq:3d model} as *deterministic* dynamical models for the sake of clarity.
In practice, we will consider a stochastic contribution to these dynamics.
This way, instead of eq. \eqref{eq:3d model}, we consider the problem of harvesting the following stochastic model:
<!--
\begin{align}
  \label{eq:may stoch}
  \Delta X_t
  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t + \eta_t,
\end{align}
where $\{\eta_t\}_t$ are independent Gaussian variables with zero mean.
The variance $\sigma^2$ of $\eta_t$ is an additional parameter of the model.

Similarly, we use the following dynamical model in place of eq. \eqref{eq:3d model}-->
\begin{align}
  \label{eq:3d model stoch}
  \begin{split}
    \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t + \eta_{X,t},\\
    \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t + \eta_{Y,t},\\
    \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t + \eta_{Z,t}.
\end{split}
\end{align}
Here, $\{\eta_{i,t}\}$ is a set of independent Gaussian variables centered at zero with variances, respectively, $\sigma_X^2$, $\sigma_Y^2$ and $\sigma_Z^2$.

# Reinforcement learning {#sec:RL}

Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
That is, the agent takes actions within the environment.
As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent. (See Fig. 1 in [@rl-intro] for a conceptual description of reinforcement learning algorithms.)
The rewards encode the agent's goal.
The main task of any RL algorithm is then to maximize the cumulative reward received. 
This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.

<!-- The *environment* is commonly a computer simulation, although it sometimes can be a real world system.
The *agent*, on the other hand, is a computer program which interacts with the environment. -->

For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $N_t=(X_t,Y_t,Z_t)$.
At each time step, the agent observes the state $S$ and harvests $X$,
\begin{align*}
  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
\end{align*}
where the fish mortality-from-harvest $M_X=M_X(N_t)$ is the agent's action at time $t$. 
This secures a reward of $M_X(N_t)X$.
Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.

## Mathematical framework for RL

The RL environment can be formally described as a discrete time *partially observable Markov decision process (POMDP)*.
This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (henceforth MDPs for short).
An MDP may be defined by the following data:

- $\mathcal{S}$: *state space*, the set of states of the environment,

- $\mathcal{A}$: *action space*, the set of actions which the agent may choose from,

- $T(N_{t+1}|N_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system (where $N_i\in\mathcal{S}$ are states of the environment),

- $r(N_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $N_t$,

- $d(N_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,

- $\gamma\in[0,1]$: *discount factor*.

At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | N_t)$.
In return, it receives a discounted reward $\gamma^t r(a_t, N_t)$.
The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.

With any fixed policy function, the agent will traverse a path 
$\tau=(N_0,\ a_0,\ N_1,\ a_1\ \dots,\ N_{t_{\text{fin.}}})$ 
sampled randomly from the distribution
\begin{align*}
  p_\pi(\tau) = 
  d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t ).
\end{align*}
Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
$$  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
$$
where,
$$
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t),
$$
is the cumulative reward of path $\tau$.
The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.

## Deep Reinforcement Learning

The policy function $\pi$ often lives in a high or even infinite-dimensional space.
This makes it unfeasible to directly optimize $\pi$.
In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.\footnote{
Policies are, in general, functions from state space to policy space.
In our case, these are real-valued functions on a cube $\pi:[0,2]^{\times 3}\to \mathbb{R}_+$.
The space of all such functions is highly singular, spanning a \emph{non-separable Hilbert space}.
Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,2]^{\times3})$.
One way to avoid optimizing over an infinite dimensional ambient space is to discretize state space into a set of bins.
This approach runs into tractability problems: First, the dimension of policy space scales exponentially with the number of species.
Second, even for a fixed number of species (e.g., 3), the dimension optimized over can be prohibitively large---for example if one uses 1000 bins for each population in a three-species model, the overall number of parameters being optimized over is $10^9$.
Neural networks with much smaller number of parameters, on the other hand, can be quite expressive and sufficient to find a rather good (if not optimal) policy function.
}
<!-- ML: I am not following this line of thought here. I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.

FM: my take here is that the space of functions from states to actions is in principle infinite-dimensional or at least very high dimensional. (e.g. exponentially high dimensional in terms of the number of degrees of freedom in state space, for the case of discretized state spaces).
By restricting our ansatz to neural networks of a fixed depth/breadth, we're restricting ourselves to a finite dimensional (or, in the discrete case, potentially much lower dimensional) manifold of policies over which gradient descent may be performed much more efficiently.
E.g. the dimensionality of the space we're searching for is determined by the size of the neural network, it's finite dimensional for continuous state spaces, and its dimension is not necessarily exponential in the number of species.
-->
Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks. (See Fig. 1 and App. A in [@rl-intro] for a conceptual introduction to the use of reinforcement learning in the context of conservation decision making.)
<!--(See Fig. \ref{tbd}.)-->
<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->

We will focus on deep reinforcement learning throughout this paper.
Within the DRL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
(See, e.g., [@franccois2018introduction] for an overview of different optimization schemes used in DRL.)

## Model-free reinforcement learning

Within control theory, the classical setup is one where we use as much information from the model as possible in order to derive an optimal solution.
Here, one may find a vast literature on model-based methods to attain optimal, or near-optimal, control (see, e.g., [@zhang2019near; @sethi2019optimal; @anderson2007optimal]).

The classical sustainable fishery management approaches summarized in Text Box 1, for instance, are model-based controls.
As we saw there, these controls may run into trouble in the case where there are inaccuracies in the model parameter estimates.

There are many situations, however, in which the exact model of the system is not known or not tractable.
This is a standard situation in ecology: mathematical models capture the most prominent aspects of the ecosystem's dynamics, while ignoring or summarizing most of its complexity.
In this case, it is clear, model-based controls run a grave danger of mismanaging the system.

Reinforcement learning, on the other hand, can provide a model-free approach to control theory.
While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
This provides more flexibility to use RL in instances where the model of the system is not accurately known.
In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances [@mbpo].
(For recent surveys of model-based reinforcement learning, which we do not focus on here, see [@moerland2023model; @polydoros2017survey].)

This context provides a motivation for this paper. 
Indeed, models for ecosystem dynamics are only ever approximate and incomplete descriptions of reality. 
This way, it is plausible that model-free RL controls outperform currently used model-based controls in ecological management problems.

# Methods

\textbf{Training a DRL agent.}
We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
Each episode in the training and in the evaluation stages was at most 200 time steps long.
Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
\begin{align}
  \label{eq:thresh}
  X_t \leq X_{\text{thresh.}},\quad
  Y_t \leq Y_{\text{thresh.}},\quad
  \text{or,}\quad
  Z \leq Z_{\text{thresh.}}.
\end{align}
The reward function had two componenets:
First, at each time step the harvest $h_t$ was received as a reward.
Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.

Thus, we normalized state space (to be contained in a cube of length 2) and engaged in reward shaping in order to optimize our agent's learning.
We did not need to tune hyper-parameters for this goal, however.

We used the Ray framework\footnote{\url{https://docs.ray.io/}} for training, in particular we used a Ray PPOConfig object to build the algorithm using the default values for all hyper-parameters.
The agent was trained for 200 iterations of the PPO optimization step.

\textbf{Tuning the constant mortality strategy.}
A grid of 100 mortality rates was laid in the interval $[0, 0.1]$.
For each one of these rates, say $M$, we simulated 50 episodes based on \eqref{eq:3d model stoch}:
At each time step the state $(X_t, Y_t, Z_t)$ was observed, a harvest of $MX_t$ was collected and then the system evolved according to its natural dynamics \eqref{eq:3d model stoch}.
The optimal mortality rate was the value $M^*$ for which the mean episode reward was maximal.
We additionally evaluated a more conservative strategy where a mortality of $0.8 M^*$ instead is used.

The result of the tuning procedure is shown in Fig. \ref{fig:tuning}. 
There, we see that the optimal value $M^*$ lies in an area with high episode reward variance.
We will see in Fig. \ref{fig:rew_len_fig} that a likely candidate for this variance is the presence of episodes with early endings.
Thus, the value $0.8 M^*$ supports a smaller variance, and correspondingly -- as Fig. \ref{fig:rew_len_fig} shows -- this policy leads to a much smaller probability of a near-extinction event.

@@

This time series is representative of the patterns observed in the data (see App. A.1).
The episode reward and episode length statistics are shown in Fig. \ref{fig:rew_len_fig}.
There, we highlight two things:
First, that the tendency towards early episode ends is fixed by the conservative approach of fishing at 80% of the MSY mortality rate.
(In App. A.2 we discuss and evaluate other policies using a fixed-percentage of the MSY mortality rate.)
Second, that while the conservative 80% approach has a stable behavior, it consistently produces smaller harvests than a DRL generated policy.

@@

# Discussion

>>>>

---
output: github_document
---

```{r setup}
library(tidyverse)
library(patchwork)
library(RColorBrewer)
```

<!-- Data sources -->
```{r}
# for best const mortality and best const escapement
msy_optimizer <- function(df) {
  df |> 
  group_by(action, rep) |>
  filter(t == max(t)) |> 
  group_by(action) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}
escapement_optimizer <- function(df) {
  df |> 
  group_by(escapement, rep) |>
  filter(t == max(t)) |> 
  group_by(escapement) |>
  summarise(mean_reward = mean(reward), sd = sd(reward)) |> 
  filter(mean_reward == max(mean_reward))
}

# Collect data (outsource to separate R script later TBD)
msy_df <- read_csv("../data/msy.csv.xz")
best_action <- msy_optimizer(msy_df)$action
escapement_df <- read_csv("../data/escapement.csv.xz", show_col_types = FALSE)
best_e <- escapement_optimizer(escapement_df)$escapement

ppo_df <- read_csv("../data/PPO220.csv.xz")

opt_escapement <- escapement_df |> filter(escapement == best_e)
```

```{r}
# set up for plots:

@@

@@

@@

<!-- Appendices -->
